```{r}
#| label: setup
#| include: false

source(here::here("R/quarto-setup.R"))
```

<!-- badges: start -->
[![Project Status: Inactive – The project has reached a stable, usable state but is no longer being actively developed; support/maintenance will be provided as time allows.](https://www.repostatus.org/badges/latest/inactive.svg)](https://www.repostatus.org/#inactive)
[![License: MIT](https://img.shields.io/badge/license-MIT-green)](https://choosealicense.com/licenses/mit/)
<!-- badges: end -->

## Overview

This document focuses on demonstrating the [Lorenz system](https://en.wikipedia.org/wiki/Lorenz_system), originally introduced by Edward N. Lorenz in his seminal [-@lorenz1963] paper. The Lorenz system comprises three coupled, nonlinear ordinary differential equations that model atmospheric convection, effectively illustrating the chaotic nature of weather patterns.

The dynamics of the model are represented by the following set of first-order, nonlinear differential equations:

$$
\begin{aligned}
\frac{dx}{dt} &= \sigma(y - x), \\
\frac{dy}{dt} &= x(\rho - z) - y \\
\frac{dz}{dt} &= xy - \beta z
\end{aligned}
$$
In these equations:

- $x$ represents the rate of convection;
- $y$ denotes the horizontal temperature variation;
- $z$ indicates the vertical temperature variation;
- $\sigma$, $\rho$, and $\beta$ are system parameters corresponding to the [Prandtl number](https://en.wikipedia.org/wiki/Prandtl_number), [Rayleigh number](https://en.wikipedia.org/wiki/Rayleigh_number), and specific physical dimensions of the fluid layer.

To learn more about the Lorenz system, see @lorenz2008.

## Setting up the environment

```{r}
#| eval: false
#| output: false

library(checkmate, quietly = TRUE)
library(deSolve, quietly = TRUE)
library(dplyr, quietly = TRUE)
library(ggplot2, quietly = TRUE)
library(gg3D, quietly = TRUE) # remotes::install_github("AckerDWM/gg3D")
library(latex2exp, quietly = TRUE)
library(magrittr, quietly = TRUE)
```

```{r}
#| echo: false

library(gg3D, quietly = TRUE)
```

## Numerical solution of the equations

```{r}
lorenz_system <- function(
    x = 1, 
    y = 1, 
    z = 1, 
    sigma = 10, 
    rho = 28, 
    beta = 8 / 3, 
    from = 0, 
    to = 100,
    by = 0.01
  ) {
  checkmate::assert_number(x)
  checkmate::assert_number(y)
  checkmate::assert_number(z)
  checkmate::assert_number(sigma)
  checkmate::assert_number(rho)
  checkmate::assert_number(beta)
  checkmate::assert_number(from, lower = 0)
  checkmate::assert_number(to, lower = from)
  checkmate::assert_number(by, lower = 0)
  
  fun <- function (t, y, parms) {
    list2env(as.list(y), envir = environment())
    list2env(as.list(parms), envir = environment())
    
    list(
      c(
        dx = sigma * (y - x),
        dy = x * (rho - z) - y,
        dz = (x * y) - (beta * z)
      )
    )
  }
  
  initial_values <- c(x = x, y = y, z = z)
  parameters <- list(sigma = sigma, rho = rho, beta = beta)
  time <- seq(from = from, to = to, by = by)
  
  data <- 
    deSolve::ode(
      y = initial_values,
      times = time, 
      func = fun,
      parms = parameters
    ) |>
    dplyr::as_tibble() |>
    dplyr::mutate(dplyr::across(dplyr::everything(), ~ as.numeric(.x)))
  
  list(
    data = data,
    initial_values = as.list(initial_values),
    parameters = as.list(parameters)
  ) |>
  invisible()
}
```

```{r}
lorenz_system() |> magrittr::extract2("data")
```

## Plotting system dynamics

```{r}
plot_system_dynamics <- function(
    x = 1, 
    y = 1, 
    z = 1, 
    sigma = 10, 
    rho = 28, 
    beta = 8 / 3, 
    from = 0, 
    to = 100,
    by = 0.01
  ) {
  checkmate::assert_number(x)
  checkmate::assert_number(y)
  checkmate::assert_number(z)
  checkmate::assert_number(sigma)
  checkmate::assert_number(rho)
  checkmate::assert_number(beta)
  checkmate::assert_number(from, lower = 0)
  checkmate::assert_number(to, lower = from)
  checkmate::assert_number(by, lower = 0)
  
  lorenz_system(x, y, z, sigma, rho, beta, from, to, by) |> 
    list2env(envir = environment())
  
  plot <- 
    data |>
    ggplot2::ggplot(ggplot2::aes(x = time)) +
    ggplot2::geom_line(
      ggplot2::aes(y = y, color = "Horizontal temperature variation (Y)"),
      linewidth = 0.5,
      alpha = 0.75
    ) +
    ggplot2::geom_line(
      ggplot2::aes(y = z, color = "Vertical temperature variation (Z)"),
      linewidth = 0.5,
      alpha = 0.75
    ) +
    ggplot2::geom_line(
      ggplot2::aes(y = x, color = "Rate of convection (X)"),
      linewidth = 0.5,
      alpha = 0.75
    ) +
    ggplot2::labs(
      title = "Lorenz System Dynamics",
      subtitle = latex2exp::TeX(
        paste0(
          "$X_0$ = ", x, " | ",
          "$Y_0$ = ", y, " | ",
          "$Z_0$ = ", z, " | ",
          "$\\sigma$ = ", round(sigma, 2), " | ",
          "$\\rho$ = ", round(rho, 2), " | ",
          "$\\beta$ = ", round(beta, 2)
        ),
      ),
      x = "Time", 
      y = "Values",
      color = ggplot2::element_blank()
    ) +
    ggplot2::scale_color_manual(
      breaks = c(
        "Rate of convection (X)", 
        "Horizontal temperature variation (Y)", 
        "Vertical temperature variation (Z)"
      ),
      values = c("blue", "red", "black")
    )
  
  print(plot)
  
  invisible()
}
```

```{r}
plot_system_dynamics()
```

## Phase space visualization

```{r}
plot_phase_space <- function(
    x = 1, 
    y = 1, 
    z = 1, 
    sigma = 10, 
    rho = 28, 
    beta = 8 / 3, 
    from = 0, 
    to = 100,
    by = 0.001,
    theta = 180,
    phi = 0
  ) {
  checkmate::assert_number(x)
  checkmate::assert_number(y)
  checkmate::assert_number(z)
  checkmate::assert_number(sigma)
  checkmate::assert_number(rho)
  checkmate::assert_number(beta)
  checkmate::assert_number(from, lower = 0)
  checkmate::assert_number(to, lower = from)
  checkmate::assert_number(by, lower = 0)
  
  lorenz_system(x, y, z, sigma, rho, beta, from, to, by) |> 
    list2env(envir = environment())

  plot <-
    data |>
    ggplot2::ggplot(ggplot2::aes(x = x, y = y, z = z, colour = time)) +
    gg3D::stat_3D(theta = theta, phi = phi, geom = "path") +
    theme_void() +
    theme(legend.position = "none")
  
  print(plot)
  
  invisible()
}
```

```{r}
plot_phase_space()
```

## References {.unnumbered}

::: {#refs}
:::
